---
title: "Exploratory Analysis in R"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installing the dplyr package


```{r}
library(dplyr)

```

### Overview of data


we will use a simple case study to demonstrate the kinds of simple graphs that can be useful in exploratory analyses. The data we will be using come from the U.S. Environmental Protection Agency (EPA), which is the U.S. government agency that sets national air quality standards for outdoor air pollution. One of the national ambient air quality standards in the U.S. concerns the long-term average level of fine particle pollution, also referred to as PM2.5. Here, the standard says that the “annual mean, averaged over 3 years” cannot exceed 12 micrograms per cubic meter. Data on daily PM2.5 are available from the U.S. EPA web site, or specifically, the EPA Air Quality System web site.

One key question we are interested is: Are there any counties in the U.S. that exceed the national standard for fine particle pollution? This question has important consequences because counties that are found to be in violation of the national standards can face serious legal consequences. In particular, states that have counties in violation of the standards are required to create a State Implementation Plan (SIP) that shows how those counties will come within the national standards within a given period of time.
```{r}
class <- c("numeric", "character", "factor", "numeric", "numeric")
pollution <- read.csv("avgpm25.csv", colClasses = class)
head(pollution)
str(pollution)
```
Each row contains the 5-digit code indicating the county (fips), the region of the country in which the county resides, the longitude and latitude of the centroid for that county, and the average PM2.5 level.

## Simple Summaries : One Dimension
For one dimensional summarize, there are number of options in R.
<ul>
<li>Five-number summary: This gives the minimum, 25th percentile, median, 75th percentile, maximum of the data and is quick check on the distribution of the data (see the fivenum()) </li>
<li>Boxplots: Boxplots are a visual representation of the five-number summary plus a bit more information. In particular, boxplots commonly plot outliers that go beyond the bulk of the data. This is implemented via the boxplot() function</li>
<li>Barplot: Barplots are useful for visualizing categorical data, with the number of entries for each category being proportional to the height of the bar. Think “pie chart” but actually useful. The barplot can be made with the barplot() function.</li>
<li>Histograms: Histograms show the complete empirical distribution of the data, beyond the five data points shown by the boxplots. Here, you can easily check skewwness of the data, symmetry, multi-modality, and other features. The hist() function makes a histogram, and a handy function to go with it sometimes is the rug() function.</li>
<li>Density plot: The density() function computes a non-parametric estimate of the distribution of a variables</li>
</ul>

### Five number summary
A five number summary can be computed with the fivenum() function, which takes a vector of numbers as input. Here, we compute a five-number summary of the PM2.5 data in the pollution dataset. 
```{r}
fivenum(pollution$pm25)
```
We can see that the median across all the counties in the dataset is about 10 micrograms per cubic meter.

For interactive work, it’s often a bit nice to use the summary() function, which has a default method for numeric vectors.

```{r}
summary(pollution$pm25)
```
### Boxplot
Here’s a quick boxplot of the PM2.5 data. Note that in a boxplot, the “whiskers” that stick out above and below the box have a length of 1.5 times the inter-quartile range, or IQR, which is simply the distance from the bottom of the box to the top of the box. Anything beyond the whiskers is marked as an “outlier” and is plotted separately as an individual point.
```{r}
boxplot(pollution$pm25, col = "blue")
```
From the boxplot, we can see that there are a few points on both the high and the low end that appear to be outliers according to the boxplot() algorithm. These points migth be worth looking at individually.

From the plot, it appears that the high points are all above the level of 15, so we can take a look at those data points directly. Note that although the current national ambient air quality standard is **12** micrograms per cubic meter, it used to be **15**.

```{r}
library(maps)
map("county", "california")
with(filter(pollution, pm25 > 15), points(longitude, latitude))
```
At this point, we might decide to follow up on these counties, or ignore them if we are interested in other features. Since these counties appear to have very high levels, relative to the distribution of levels in the other counties in the U.S., they may be worth following up on if we are interested in describing counties that are potentially in violation of the standards.

ote that the plot/map above is not very pretty, but it was made quickly and it gave us a sense of where these outlying counties were located and conveyed enough information to help decide if we an to follow up or not.

## Histogram
```{r}

```
A histogram is useful to look at when we want to see more detail on the full distribution of the data. The boxplot is quick and handy, but fundamentally only gives us a bit of information.

Here is a histogram of the PM2.5 annual average data.

```{r}
hist(pollution$pm25, col = "green")
```
This distribution is interesting because there appears to be a high concentration of counties in the neighborhood of 9 to 12 micrograms per cubic meter. We can get a little more detail of we use the rug() function to show us the actual data points.

```{r}
hist(pollution$pm25, col = "green")
rug(pollution$pm25)
```
The large cluster of data points in the 9 to 12 range is perhaps not surprising in this context. It’s not uncommon to observe this behavior in situations where we have a strict limit imposed at a certain level. Note that there are still quite a few counties above the level of 12, which may be worth investigating.

The hist() function has a default algorithm for determining the number of bars to use in the histogram based on the density of the data (see ?nclass.Sturges). However, you can override the default option by setting the breaks argument to something else. Here, we use more bars to try to get more detail.

```{r}
hist(pollution$pm25, col = "green", breaks = 100)
rug(pollution$pm25)
```
Now we see that there’s a rather large spike 9 micrograms per cubic meter. It’s not immediately clear why, but again, it might be worth following up on.
Once we start seeing interesting features in our data, it’s often useful to lay down annotations on our plots as reference points are markers. For example in our boxplot above, we might want to draw a horizontal line at 12 where the national standard is.

```{r}
 boxplot(pollution$pm25, col = "blue")
abline(h = 12)
```
We can see that a reasonable portion of the distribution as displayed by the boxplot is above the line (i.e. potentially in violation of the standard).

While the boxplot gives a sense, the histogram might be better suited to visualizing the data here. In the plot below, we see the histogram and draw two lines, one at the median of the data and one at 12, the level of the standard.
```{r}
hist(pollution$pm25, col = "green")
abline(v = 12, lwd = 2)
abline(v = median(pollution$pm25), col = "magenta", lwd = 4)
```
Note that for the vertical lines, we can use both color (col) and the line width (lwd) to indicate different components of information.

### Barplot
The barplot is useful for summarizing categorical data. Here we have one categorical variable, the region in which a county resides (east or west). We can see how many western and eastern counties there are with barplot(). We use the table() function to do the actual tabulation of how many counties there are in each region.
```{r}
table(pollution$region) %>% barplot(col = "wheat")
```
We can see quite clearly that there are many more counties in the eastern U.S. in this dataset than in the western U.S.

## Simple Summaries: Two Dimensions and Beyond
So far we’ve covered some of the main tools used to summarize one dimensional data. For investigating data in two dimensions and beyond, there is an array of additional tools. Some of the key approaches are

<ul>
<li>Multiple or overlayed 1-D plots (Lattice/ggplot2): Using multiple boxplots or multiple histograms can be useful for seeing the relationship between two variables, especially when on is naturally categorical.</li>
<li>Scatterplots: Scatterplots are the natural tool for visualizing two continuous variables. Transformations of the variables (e.g. log or square-root transformation) may be necessary for effective visualization.</li>
<li>Smooth scatterplots: Similar in concept to scatterplots but rather plots a 2-D histogram of the data. Can be useful for scatterplots that may contain many many data points.</li>
<li>Overlayed or multiple 2-D plots; conditioning plots (coplots): A conditioning plot, or coplot, shows the relationship between two variables as a third (or more) variable changes. For example, you might want to see how the relationship between air pollution levels and mortality changes with the season of the year. Here, air pollution and mortality are the two primary variables and season is the third variable varying in the background.</li>
<li>Use color, size, shape to add dimensions: Plotting points with different colors or shapes is useful for indicating a third dimension, where different colors can indicate different categories or ranges of something. Plotting symbols with different sizes can also achieve the same effect when the third dimension is continuous.</li>
<li>Spinning/interactive plots: Spinning plots can be used to simulate 3-D plots by allowing the user to essentially quickly cycle through many different 2-D projections so that the plot feels 3-D. These are sometimes helpful to capture unusual structure in the data, but I rarely use them.</li>
<li>Actual 3-D plots (not that useful): Actual 3-D plots (for example, requiring 3-D glasses) are relatively few and far between and are generally impractical for communicating to a large audience. Of course, this may change in the future with improvements in technology…</li>
</ul>

### Multiple Boxplots
One of the simplest ways to show the relationship between two variables (in this case, one categorical and one continuous) is to show side-by-side boxplots. Using the pollution data described above, we can show the difference in PM2.5 levels between the eastern and western parts of the U.S. with the boxplot() function.
```{r}
boxplot(pm25 ~ region, data = pollution, col = "red")

```
### Multiple Histogram
It can sometimes be useful to plot multiple histograms, much like with side-by-side boxplots, to see changes in the shape of the distribution of a variable across different categories. However, the number of histograms that you can effectively put on a page is limited.
Here is the distribution of PM2.5 in the eastern and western regions of the U.S.
```{r}
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))
hist(subset(pollution, region == "east")$pm25, col = "green")
hist(subset(pollution, region == "west")$pm25, col = "green")
```
We can see here that the PM2.5 in the western U.S. tends to be right-skewed with some outlying counties with very high levels. The PM2.5 in the east tends to be left skewed some counties having very low levels.

### Scatterplots
For continuous variables, the most common visualization technique is the scatterplot, which simply maps each variable to an x- or y-axis coordinate. Here is a scatterplot of latitude and PM2.5, which can be made with the plot() function.
```{r}
with(pollution, plot(latitude, pm25))
abline(h = 12, lwd = 2, lty = 2)
```
As we go from south to north in the U.S., we can see that the highest levels of PM2.5 tend to be in the middle region of the country.

### Scatterplot - using Color
If we wanted to add a third dimension to the scatterplot above, say the region variable indicating east and west, we could use color to highlight that dimension. Here we color the circles in the plot to indicate east (black) or west (red).
```{r}
with(pollution, plot(latitude, pm25, col = region))
abline(h = 12, lwd = 2, lty = 2)

```
It may be confusing at first to figure out which color gets mapped to which region. We can find out by looking directly at the levels of the region variable.
Here we see that the first level is “east” and the second level is “west”. So the color for “east” will get mapped to 1 and the color for “west” will get mapped to 2. For plotting functions, col = 1 is black (the default color) and col = 2 is red


```{r}
levels(pollution$region)
```
### Multiple Scatterplot
Using multiple scatterplots can be necessary when overlaying points with different colors or shapes is confusing (sometimes because of the volume of data). Separating the plots out can sometimes make visualization easier.

```{r}
par(mfrow = c(1, 2), mar = c(5, 4, 2, 1))
with(subset(pollution, region == "west"), plot(latitude, pm25, main = "West"))
with(subset(pollution, region == "east"), plot(latitude, pm25, main = "East"))
```

These kinds of plots, sometimes called panel plots, are generally easier to make with either the lattice or ggplot2 system

```{r}
## Lattice
library(lattice)
xyplot(pm25 ~ latitude | region, data = pollution)
 
> ## ggplot2
library(ggplot2)
qplot(latitude, pm25, data = pollution, facets = . ~ region)
```

